---
title: "Exam for applicants – GRAS - Hassam Ali"
output: html_notebook
---


Task -1: Analysis of tree cover loss from 2000-2025 (based on Global Forest Change data)
```{r}
library(sf)
library(terra)
library(tmap)
library(dplyr)
```


Read point shapefile and reproject it to EPSG:5387
```{r}
pt <- st_read("D:/GRAS/study_point.shp")
pt_utm <- st_transform(pt, 5387)

#display crs info for a quick check
st_crs(pt_utm)

```


Create a 25km buffer and save it
```{r}
#increase the 'nQuadSegs' to get a more smooth and a very accurate buffer
buffer <- st_buffer(pt_utm, dist = 25000, nQuadSegs = 10000)

st_write(buffer, "D:/GRAS/study_area.shp", driver = "ESRI Shapefile", append=FALSE)
```


Read deforestation raster
```{r}
rast <- rast("D:/GRAS/Hansen_GFC-2025-v1.13_lossyear_00N_080W.tif")

rast
```


Reproject, Crop & Clip the deforestation Raster
```{r}
# 1. Convert sf -> SpatVector
buffer_v <- vect(buffer)

# 2. Reproject buffer to raster CRS
temp_buffer <- project(buffer_v, crs(rast))

# 3. Crop raster in its native CRS
rast_crop <- crop(rast, temp_buffer)

# 4. Reproject cropped raster to EPSG:5387
rast_utm  <- project(rast_crop, "EPSG:5387", method="near")

# 5. Reproject buffer to EPSG:5387
buffer_utm <- project(buffer_v, "EPSG:5387")

# 6. Mask
rast_clip <- mask(rast_utm, buffer_utm)


```


Save the clipped Raster
```{r}
writeRaster(rast_clip, "D:/GRAS/Deforest_clip.tif", overwrite = TRUE)
```
```{r}
Sys.unsetenv("PROJ_LIB")

```


Remove the 0 value where no forest loss occurred
```{r}
rast_clip[rast_clip == 0] <- NA

plot(rast_clip)
```


Vectorize the raster without the no-data values and save as a shapefile
```{r}
rast_poly <- as.polygons(
  rast_clip,
  dissolve = TRUE,
  na.rm = TRUE
)

rast_poly_sf <- st_as_sf(rast_poly) #this time convert SpatVector -> sf
st_write(rast_poly_sf, "D:/GRAS/deforestation.shp", append=FALSE)
```


Check the issue with the field name
```{r}
shp <- st_read("D:/GRAS/deforestation.shp")
names(shp)
```


```{r}
head(shp)
```


Year went into H_GFC.2 , Rename it to year
```{r}
names(shp)[names(shp) == "H_GFC.2"] <- "Year"
```


Save the updated file
```{r}
st_write(shp, "D:/GRAS/deforestation.shp", delete_dsn = TRUE)
```


Static Plot
```{r}
final_shp <- st_read("D:/GRAS/deforestation.shp")

tmap_mode("plot")

tm_shape(final_shp) +
  tm_polygons(
    col = "Year",
    palette = "Spectral",
    style = "cat",
    title = "Year-wise Deforestation",
    border.col = NA,
    showNA = FALSE,
    drop.levels = TRUE
    
  ) +
  tm_shape(buffer) +
  tm_borders(col = "black", lwd = 2)
```

Interactive map with additional layers and custom color Platte
```{r}
final_shp <- st_read("D:/GRAS/deforestation.shp")
buffer <- st_read("D:/GRAS/study_area.shp")
#Custom Color Platte created with the help of ColorBrewer
cols <- colorRampPalette(
  c(
  "#5C0303", "#8B1101", "#CE2D04", "#FF2B05",
  "#F97C1D", "#FEB130", "#CCC53A", "#DBE236",
  "#C8F735", "#8BFF4B", "#52FA7A", "#22EBAA",
  "#1AD4D0", "#2FB2F4", "#458CFD", "#4666DD",
  "#3F3E9C", "#150183"
  )
)(25)


#Read /load the layers
places <- st_read("D:/GRAS/Places.shp")
prot_area <- st_read("D:/GRAS/Protected Areas.shp")
rivers <- st_read("D:/GRAS/Rivers & Streams.shp")
tracks <- st_read("D:/GRAS/Roads and Tracks.shp")

rivers      <- st_transform(rivers,      st_crs(final_shp))
tracks      <- st_transform(tracks,      st_crs(final_shp))
places      <- st_transform(places,      st_crs(final_shp))
prot_area   <- st_transform(prot_area,   st_crs(final_shp))
buffer      <- st_transform(buffer,      st_crs(final_shp))

tmap_mode("view")
tm_basemap(
  server = "Esri.WorldTopoMap"
) +
tm_shape(final_shp) +
  tm_polygons(
    col = "Year",
    palette = cols,
    style = "cat",
    title = "Year-wise Deforestation",
    border.col = NA,
    showNA = FALSE,
    drop.levels = TRUE
  ) +

  # Rivers in blue
  tm_shape(rivers) +
  tm_lines(col = "#0087ff", lwd = 2) +

  # Tracks in dark grey
  tm_shape(tracks) +
  tm_lines(col = "grey20", lwd = 1.5) +

  # Protected area with dotted fill
  tm_shape(prot_area) +
  tm_polygons(
    col = NA,
    border.col = "darkgreen",
    lwd = 1.5,
    fill = "darkgreen",
    fill_alpha = 0.15,
    lty = "dotted"
  ) +

  # Places as points with labels
  tm_shape(places) +
  tm_symbols(col = "red", size = 0.1) +
  tm_text("name", size = 0.9, col = "black", ymod = -0.5) +

  # Buffer boundary
  tm_shape(buffer) +
  tm_borders(col = "red", lwd = 0.5, lty = "dashed")

```


Calculate the area
```{r}
final_shp$area_m2 <- st_area(final_shp)

final_shp$area_ha <- as.numeric(final_shp$area_m2) / 10000
```


check
```{r}
head(final_shp)
```


Add years
```{r}
final_shp$year <- final_shp$Year + 2000
```


check
```{r}
head(final_shp)
```


Summarize by year
```{r}
annual_area <- final_shp %>%
  group_by(year) %>%
  summarise(
    area_ha = sum(area_ha, na.rm = TRUE)
  )
```



check
```{r}
head(annual_area)
```



area in 2021
```{r}
area_2021 <- annual_area %>%
  filter(year == 2021)
```


area in 2008-2025
```{r}
total_2008_2025 <- annual_area %>%
  filter(year >= 2008 & year <= 2025) %>%
  summarise(area_ha = sum(area_ha))
```



area 2021-2025
```{r}
total_2021_2025 <- annual_area %>%
  filter(year >= 2021 & year <= 2025) %>%
  summarise(area_ha = sum(area_ha))
```



Create the final table
```{r}
results <- data.frame(
  Period = c(
    "Deforestation in 2021",
    "Total 2008-2025",
    "Total 2021-2025"
  ),
  Area_ha = c(
    area_2021$area_ha,
    total_2008_2025$area_ha,
    total_2021_2025$area_ha
  )
)

results
```

